Local Type Inference for Polarised System F with Existentials

### 2023-10-31 oopsla24/tex/01-introduction.tex
fooo

There has been a limited amount of work on support for existential types as a first-class concept. In an unpublished draft, \citet{leijen06} studied an extension of Damas-Milner inference in which type schemes contain alternating universal and existential quantifiers. Quantifiers still range over monotypes, and higher-rank polymorphism is not permitted. More recently, \citet{dk19} studed type inference for existential types in the context of GADT type inference, which, while still predicative, supported higher-rank types (i.e., quantifiers can occur anywhere inside of a type scheme). \citet{existential-crisis} propose a system which is intermediate between prenex and higher-rank polymorphism, but which

All of these papers are restricted to \emph{predicative} quantification, whereas existential types in full System F are \emph{impredicative} -- that is, quantifiers can be instantiated with

\begin{itemize}
\item There has been very little work on type inference for existentials. 
\item The state of the art in languages like Ocaml and Haskell is the
  Odersky-L\"{a}ufer algorithm, which makes existentials a
  second-class construct tied to datatypes.
\item More recently, there's the existential crisis paper, which
  showed how to extend ML-style inference with some support for
  existentials. However, it is (a) predicative-only, (b) restricted
  for $\forall\exists$ types, and (c) does not have a declarative
  specification. (In fact, the authors comment on the need for such.)
\item In this paper, we lift all of these restrictions. We give a
  local type inference algorithm which supports first-class
  existentials, for impredicative types, with no restrictions on
  nested quantifers, and we have declarative specification we prove
  our algorithm sound and complete with respect to.
\item Developing the algorithm required us to break some of the
  fundamental invariants of HM-style typin, and as a result, our
  algorithm needs to mix unification and anti-unification, and works
  over a call-by-push-value metalanguage to control how quantifiers
  are instantiated and things like function arities.
\item Unfortunately, local type inference does not infer types anywhere
  nearly as comprehensively as full Damas-Milner type inference. We
  do guarantee that all type applications (for $\forall$-elimination) 
  and all packs (for $\exists$-introdution) are inferred, but functions
  and some let-bindings still need annotations on their arguments.
\item The original local type inference paper combined local type inference
  with bidirectional typechecking to minimize the number of needed
  annotations, but we show how existential types complicate the integration
  of bidirectionality with local type inference, and we explore the design space
  to show how the same scheme could be applied to work with different
  type systems. 
\end{itemize}

>>>>

There has been a limited amount of work on support for existential types as a first-class concept. In an unpublished draft, \citet{leijen06} studied an extension of Damas-Milner inference in which type schemes contain alternating universal and existential quantifiers. Quantifiers still range over monotypes, and higher-rank polymorphism is not permitted. More recently, \citet{dk19} studed type inference for existential types in the context of GADT type inference, which, while still predicative, supported higher-rank types (i.e., quantifiers can occur anywhere inside of a type scheme). \citet{existential-crisis} propose a system which only permits types of the form forall-exists, but which permits projective elimination in the style of ML modules.

All of these papers are restricted to \emph{predicative} quantification, where quantifiers can only be instantiated with monotypes (i.e., types without any occurences of quantifiers). However, existential types in full System F are \emph{impredicative} -- that is, quantifiers can be instantiated with arbitrary types, specifically including types containing quantifiers.

Historically, inference for impredicative quantification has been neglected, due to results by \citet{tiuryn-urzczyn-96} and \citet{chrzaszcz-98} which show that not only is full type inference for System F undecidable, but even that the subtyping relation induced by type instantiation is undecidable. However, in recent years, interest in impredicative inference has revived (for example, the work of \citet{serrano-2020}), with a focus on avoiding the undecidabilty barriers by doing \emph{partial} type inference. That is, we no longer try to do full type inference, but rather accept that the programmer will need to write annotations in cases where inference would be too difficult. Unfortunately, it is often difficult to give a specification for what the partial algorithm does -- for example, \citet{existential-crisis} observe that their algorithm lacks a declarative specification, and explain why existential types make this particularly difficult to do.

One especially well-behaved form of partial type inference is \emph{local type inference}, introduced by \citet{pierce-local-2000}. In this approach, quantifiers in a polymorphic function are instantiated using only the type information available in the arguments at each call site. While this infers fewer quantifiers than global algorithms such as Damas-Milner can, it has the twin benefits that it is easy to give a mathematical specification to, and that failures of type inference are easily explained in terms of local features of the call site.

In this paper, we extend the local type inference algorithm to a language with both universal and existential quantifiers, which can both be instantiated impredicatively. This combination of features broke a number of the invariants which traditional type inference algorithms depend on, and required us to invent new algorithms which combine both unification and (surprisingly) anti-unification.

\paragraph{Contributions} Our contributions are as follows: 
\begin{itemize}
\item We give a declarative type system (again, based on call-by-push-value) to serve as a specification of our algorithm, and we we prove our algorithm is sound and complete with respect to the declarative type system. The specification makes it easy to see that all type applications (for $\forall$-elimination) and all packs (for $\exists$-introdution) are inferred.

\item We give a local type inference algorithm which supports both first-class existential and universal quantifiers, both of which can be instantiated impredicatively. To evade the undecidability results surrounding type inference for System F, we work in a variant of call-by-push-value, which lets us formulate a subtyping relation which is still decidable.

\item Our algorithm breaks some of the fundamental invariants of
  HM-style type inference. As a result, it needs to mix unification
  and anti-unification, uses the call-by-push-value structure to
  control function arities and how quantifiers can be instantiated.

\item The original local type inference paper combined local type inference
  with bidirectional typechecking to minimize the number of needed
  annotations, but we show how existential types complicate the integration
  of bidirectionality with local type inference, and we explore the design space
  to show how the same scheme could be applied to work with different
  type systems. 
\end{itemize}
